Calculating the nonequilibrium observable by a speed-boosting technique

### 2013-10-22 writedown/accelerate/main.tex
add latest result

\newcommand{\recheck}[1]{{\color{red} #1}}
\newcommand{\redc}[1]{{\color{red} #1}}
\newcommand{\bluec}[1]{{\color{blue} #1}}
\newcommand{\vect}[1]{\textbf{\textit{#1}}}
\newcommand{\dd}{\textsf{d}}
\newcommand{\inv}{\textrm{inv}}

@@

>>>>

\newcommand{\recheck}[1]{{\color{red} #1}}
\newcommand{\redc}[1]{{\color{red} #1}}
\newcommand{\bluec}[1]{{\color{blue} #1}}
\newcommand{\vect}[1]{\textbf{\textit{#1}}}
\newcommand{\dd}{\textsf{d}}
\newcommand{\inv}{\textrm{inv}}
\newcommand{\hard}{\textrm{h}}
\newcommand{\soft}{\textrm{s}}
\newcommand{\vdw}{\textrm{vdw}}
\newcommand{\ele}{\textrm{ele}}
\newcommand{\dir}{\textrm{dir}}
\newcommand{\rec}{\textrm{rec}}
\newcommand{\corr}{\textrm{corr}}

@@

\section{The slow and fast part of the CHARMM force field}

In this study, we are using the following CHARMM27 force field~\cite{mackerell2004extending}, which can be described:
\begin{align}\nonumber
  U(\vect R) = &
  \sum_{\textrm{bonds}} K_b(b-b_0)^2\\\nonumber
  &
  +\sum_{\textrm{angles}} K_\theta (\theta - \theta_0)^2
  +\sum_{\textrm{Urey-Bradley}} K_{\textrm{UB}} (S - S_0)^2\\\nonumber
  &
  +\sum_{\textrm{dihedrals}} K_\phi( 1+\cos(n\phi - \delta)) 
  +\sum_{\textrm{impropers}} K_\omega (\omega - \omega_0)^2\\ \nonumber
  &
  +\sum_{\textrm{residues}} U_{\textrm{CMAP}}(\phi, \psi) \\\label{eqn:tmp13}
  &
  +\sum_{\textrm{non-bonded}}
  \big[\,U_{\vdw}(r_{ij}) + U_{\ele}(r_{ij})\,\big]
  % \Bigg\{
  % \eps_{ij}
  % \bigg[\bigg(\frac{R_{ij}}{r_{ij}}\bigg)^{12}
  % - 2\bigg(\frac{R_{ij}}{r_{ij}}\bigg)^{6}
  % \bigg]
  % + \frac{q_iq_j}{4\pi\eps_0\eps r_{ij}}
  % \Bigg\}
\end{align}
The terms include bond ($b$), angle ($\theta$), Urey-Bradley ($S$),
dihedral angle ($\phi$), improper angle ($\omega$), backbone torsional
correction (CMAP) and non-bonded van der Waals and electrostatic
interactions.  The van der Waals interaction is written by
\begin{align}
  U_{\vdw}(r_{ij}) = 
  \eps_{ij}
  \bigg[\bigg(\frac{R_{ij}}{r_{ij}}\bigg)^{12}
  - 2\bigg(\frac{R_{ij}}{r_{ij}}\bigg)^{6}
  \bigg]
\end{align}
The electrostatic interaction is
\begin{align}
  U_{\ele}(r_{ij}) = \frac{q_iq_j}{4\pi\eps_0\eps r_{ij}}
\end{align}
We treat the electrostatic interaction by the Particle Mesh Ewald (PME)
method~\cite{darden1993pme, essmann1995spm}, which split the electrostatic interaction into three parts,
namely the direct, the reciprocal and the correction parts.
\begin{align}
  U_{\ele}(r_{ij}) =
  U_{\dir}(r_{ij}) +
  U_{\rec}(r_{ij}) + U_{\corr},
\end{align}
where
\begin {align}
  U_{\dir} (r_{ij})& = \frac12 
  \frac{q_iq_j \textrm{erfc}(\beta {r}_{ij})}
  {{r}_{ij} } \\
  U_{\rec} (r_{ij})& = \frac1{2\pi V} \sum_{\vect m \neq 0}
  \frac{\exp(-\pi^2 m^2 / \beta^2)}{m^2}
   q_iq_j e^ {2 \pi i m \cdot r_{ij}}\\
  U_{\corr}& = -\frac\beta{\sqrt \pi} q_iq_j \delta_{ij}
\end {align}
The correction part is a constant term with respect to the position of
the atoms, so it is not calculated during
the force calculation.\\

\noindent
Among all the components of the CHARMM force field, the bond
interaction is treated as fast interaction. The angle, Urey-Bradley,
dihedral angle, improper angle, backbone torsional correction and the
reciprocal part of the electrostatic interaction are treated as slow
interactions. The van der Waals interaction has a hard core,
so it is split into soft and hard contributions in the following way:
\begin{align}
  U_{\vdw} (r) &= U^\hard_{\vdw}(r) + U^\soft_{\vdw}(r)
\end{align}
and
\begin{align}
  U^\hard_{\vdw}(r) & =  [\, 1-w_\vdw - (1 - w_\vdw) s_\vdw(r) \,] \,U_{\vdw}(r)\\
  U^\soft_{\vdw}(r) & =  [\, w_\vdw + (1 - w_\vdw) s_\vdw(r) \,] \,U_{\vdw}(r)
\end{align}
where $w_\vdw$ is a positive parameter between 0 and 1, and
$s_\vdw(r)$ is a function smoothly switches from 0 to 1:
\begin{align}
  s_\vdw(r) =
  \left\{
    \begin{aligned}
      &\frac12 - \frac12 \cos \Big(\frac{r}{r^\hard_\vdw} \pi\Big) &\quad & r < r^\hard_\vdw\\
      &1 & \quad & r \geq r^\hard_\vdw
    \end{aligned}
  \right.
\end{align}
The soft part of the van der Waals $U^\soft_{\vdw}(r)$ switches
smoothly from $w\,U_{\vdw}(r)$ to $U_{\vdw}(r)$ in the range
$[0, r^\hard_\vdw]$. Here $[0, r^\hard_\vdw]$ is the range where the the van der Waals is
treated as hard interaction, and therefore has to been reduced. The
direct part of the electrostatic interaction is split in the same way:
\begin{align}
  U_{\dir} (r) &= U^\hard_{\dir}(r) + U^\soft_{\dir}(r)
\end{align}
and
\begin{align}
  U^\hard_{\dir}(r) & =  [\, 1-w_\dir - (1 - w_\dir) s(r) \,] \,U_{\dir}(r)\\
  U^\soft_{\dir}(r) & =  [\, w_\dir + (1 - w_\dir) s(r) \,] \,U_{\dir}(r)
\end{align}
where $w_\dir$ is a positive parameter between 0 and 1 that
has the similar meaning as $w_\vdw$. \\

\noindent
In summary, the CHARMM force field is split in the following way:
\begin{align}
  U(\vect R) = U^\hard(\vect R) + U^\soft(\vect R)
\end{align}
where
\begin{align}
  U^\hard(\vect R) &=
  \sum_{\textrm{bonds}} K_b(b-b_0)^2 +
  \sum_{\textrm{non-bonded}} [\,U_{\vdw}^\hard (r_{ij}) + U_{\dir}^\hard (r_{ij})\,] \\\nonumber
  U^\soft(\vect R) &=
  +\sum_{\textrm{angles}} K_\theta (\theta - \theta_0)^2
  +\sum_{\textrm{Urey-Bradley}} K_{\textrm{UB}} (S - S_0)^2\\\nonumber
  &
  +\sum_{\textrm{dihedrals}} K_\phi( 1+\cos(n\phi - \delta)) 
  +\sum_{\textrm{impropers}} K_\omega (\omega - \omega_0)^2\\ \nonumber
  &
  +\sum_{\textrm{residues}} U_{\textrm{CMAP}}(\phi, \psi) \\
  &
  +\sum_{\textrm{non-bonded}}
  \big[\,U^\soft_{\vdw}(r_{ij})
  + U^\soft_{\dir}(r_{ij})
  + U_{\rec}(r_{ij})\,\big]  
\end{align}
Therefore, the interaction for the accelerated system is
\begin{align}
  \hat U(\vect R) = U^\hard(\vect R) + \alpha\,U^\soft(\vect R)  
\end{align}

\section{Numerical results}
\noindent
We study the alanine dipeptide under an oscillatory EF for example.
This example is exactly the same as the period $T=40$~ps case in the previous
manuscript~\cite{wang2013exploring}. The only difference is that we use
flexible bonds instead of rigid bonds in the simulations, and the corresponding
time-step is now 0.5~fs. The split parameter $\beta$ in the PME method is optimzied by
the method mentioned in Ref.~\cite{wang2010optimizing}\\

\begin{table}
  \centering
  \begin{tabular*}{0.48\textwidth}{@{\extracolsep{\fill}}lcccc}
    \hline\hline
    $\alpha$ & $w_\vdw$ & $r^\hard_\vdw$ [nm] & $w_\dir$ & $r^\hard_\vdw$ [nm] \\\hline
    2           & 1.00     & 0.0   & 1.00    & 0.0\\
    5           & 1.00     & 0.0   & 0.80    & 0.2\\
    6           & 1.00     & 0.0   & 0.67    & 0.2\\
    \hline\hline
  \end{tabular*}
  \caption{The systems tested and the parameters to split the slow and fast parts of the force.}
  \label{tab:tmp1}
\end{table}

\noindent
Till now we have tested the following three systems listed in
Tab.~\ref{tab:tmp1}. The probabilities of different conformations as a
function of time are plotted in Fig.~\ref{fig:tmp1}--\ref{fig:tmp3}.
We can see that when $\alpha = 2$, the result is consistent with the
reference simulation (no speed boosting, i.e.~$\alpha = 1$). For
bigger $\alpha = 5$ and 6, the small time scale oscillating (20~ps
that is roughly the same as half of the period of the EF) of
conformations $A_1$, $A_2$, $B_1$ and $B_2$ is not well captured by
the speed boosting algorithm. Moreover, the probability of $B_1$ is
overestimated, while the probability of $B_2$ is underestimated (see
the blue and purple lines in Fig.~\ref{fig:tmp2} and \ref{fig:tmp3}).
We plot the time dependent probability density of $\alpha$-helix
($A_1+A_2$) and $\beta$-sheet ($B_1+B_2$) conformations, on the right
plots of Fig.~\ref{fig:tmp2} and \ref{fig:tmp3}. We can see that for
$\alpha=5$, the result consistent with the reference simulation very
well, while for $\alpha=6$ the precision is slightly lower, but is
still qualitatively correct.  That means that the fine-scaled
oscillating of conformation and the probability exchange between the
two sub-conformations of the $\beta$-sheet are statistically relevant
to the hard forces that are modified during the speed boosting
simulation, while the conformation changes between the $\alpha$-helix
$\beta$-sheet and left-handed $\alpha$-helix ($C$)  are not correlated
with the hard forces.

\begin{figure}
  \centering
  \includegraphics[]{figs/fig-meta02.eps}
  \caption{$\alpha = 2$, the probability of different conformations as
    a function of time.  Red: $A_1$, green: $A_2$, blue: $B_1$,
    purple: $B_2$, cyan: $C$.  The lighter colors denote the reference
    exact dynamics of the conformations ($\alpha = 1$), while the
    darker colors denote $\alpha = 2$.}
  \label{fig:tmp1}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[]{figs/fig-meta04.eps}
  \includegraphics[]{figs/fig-meta04-1.eps}
  \caption{$\alpha = 4$, the probability of different conformations as a function of time.
    In the left plot: Red: $A_1$, green: $A_2$, blue: $B_1$, purple: $B_2$, cyan: $C$.
    In the right plot: Red: $A_1+A_2$, blue: $B_1+B_2$, cyan: $C$.
    The lighter colors denote the reference exact dynamics of the conformations ($\alpha = 1$), while the darker colors denote $\alpha = 4$.}
  \label{fig:tmp2}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[]{figs/fig-meta05.eps}
  \includegraphics[]{figs/fig-meta05-1.eps}
  \caption{$\alpha = 5$, the probability of different conformations as a function of time.
    In the left plot: Red: $A_1$, green: $A_2$, blue: $B_1$, purple: $B_2$, cyan: $C$.
    In the right plot: Red: $A_1+A_2$, blue: $B_1+B_2$, cyan: $C$.
    The lighter colors denote the reference exact dynamics of the conformations ($\alpha = 1$), while the darker colors denote $\alpha = 5$.}
  \label{fig:tmp2}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[]{figs/fig-meta08.eps}
  \includegraphics[]{figs/fig-meta08-1.eps}
  \caption{$\alpha = 6$, the probability of different conformations as a function of time.
    In the left plot: Red: $A_1$, green: $A_2$, blue: $B_1$, purple: $B_2$, cyan: $C$.
    In the right plot: Red: $A_1+A_2$, blue: $B_1+B_2$, cyan: $C$.
    The lighter colors denote the reference exact dynamics of the conformations ($\alpha = 1$), while the darker colors denote $\alpha = 6$.}
  \label{fig:tmp3}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-02-1.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-02-2.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-02-3.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-02-4.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-02-5.eps}\\
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-04-1.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-04-2.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-04-3.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-04-4.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-04-5.eps}\\
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-05-1.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-05-2.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-05-3.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-05-4.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-05-5.eps}\\
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-08-1.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-08-2.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-08-3.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-08-4.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-08-5.eps}\\
  \caption{The integrated probability fluxes. $\alpha = 2,4,5,8$.}
  \label{fig:tmp3}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-04-1-1.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-04-1-2.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-04-1-3.eps}\\
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-05-1-1.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-05-1-2.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-05-1-3.eps}\\
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-08-1-1.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-08-1-2.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-08-1-3.eps}
  \caption{The integrated probability fluxes. $\alpha = 4,5,8$.}
  \label{fig:tmp3}
\end{figure}

\section*{References}
\bibliography{ref}{}
\bibliographystyle{unsrt}
